---
title: "Cell-level representation learning"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(ggplot2)
```

The goal: turn raw, multi-batch, multi-modal single-cell data into one clean
**embedding**, a handful of coordinates per cell that mix the batches together
while keeping the biology, so you can cluster and analyze all your data as one.

This runs on the full **Liu et al.** COVID-19 dataset (RNA + ADT).

```{r libs}
library(m3)
set.seed(0)
```

## 1. Read the batches and merge

Each batch is one run: an RNA HVG matrix, an ADT matrix, and a
metadata CSV. `m3_read_h5()` reads one batch into an `m3_dataset`; `m3_concat()`
stacks them, harmonising features by intersection and stamping a `batch` column.
(Adjust `DATA` to where your Liu et al files live.)

```{r data}
DATA <- Sys.getenv("M3_LIU_DATA", "/path/to/liu_et_al")
cohorts <- lapply(1:3, function(i) m3_read_h5(
  rna      = file.path(DATA, sprintf("rna%d_hvg1000.h5", i)),
  adt      = file.path(DATA, sprintf("adt%d.h5", i)),
  metadata = file.path(DATA, sprintf("metadata%d.csv", i)),
  batch    = sprintf("B%d", i)))
data <- m3_concat(cohorts)
data
```

## 2. Build and train the model

`m3_train(...)` sets up the model from your data and fits it, telling it which
obs columns to use:

- `condition_keys`, the conditions you're comparing across (here disease group +
  age band).
- `celltype_key`, the column holding cell-type labels.
- `batch_key`, the column marking which batch each cell is from (B1/B2/B3). The
  model corrects for batch, so cells group by biology rather than by batch.
- `embedding_dim`, how many numbers describe each cell in the output (here 30).

`donor_prediction = FALSE` fits only the integration model and skips the
disease-prediction step (not needed for embeddings, and faster). `seed = 0`
makes the run reproducible. Afterwards `m3_capabilities(model)` shows which
outputs are available.

```{r train}
model <- m3_train(
  data,
  condition_keys = c("cond_group", "Age_interval"),
  celltype_key   = "mergedcelltype",
  batch_key      = "batch",
  embedding_dim  = 30L,
  donor_prediction = FALSE,
  max_epochs = 500L,
  seed = 0L
)
model
m3_capabilities(model)
```

## 3. Get the embedding

`m3_embedding(model, part = ...)` returns the trained coordinates as a
cells x embedding_dim matrix. Ask for the part you want by name:

- `"bio"`, the biological signal, with batch differences removed. This is the
  one you cluster and analyze.
- `"batch"`, the batch signal on its own, if you ever want to inspect it
  separately.

`m3_cell_metadata(model)` is the per-cell obs table, row-aligned with the
embedding.

```{r embed}
emb_bio       <- m3_embedding(model, part = "bio")
emb_intrinsic <- m3_embedding(model, part = "intrinsic")
emb_batch     <- m3_embedding(model, part = "batch")
meta <- m3_cell_metadata(model)
cat(sprintf("bio: %s | intrinsic: %s | batch: %s\n",
            paste(dim(emb_bio), collapse = "x"),
            paste(dim(emb_intrinsic), collapse = "x"),
            paste(dim(emb_batch), collapse = "x")))
```

## 4. Save the embedding + metadata

Save the embedding and its metadata. m3 returns plain matrices, so persist them
however you like, here as CSV and an `.rds`.

```{r save}
out <- file.path(tempdir(), "m3_tut1")
dir.create(out, showWarnings = FALSE)
utils::write.csv(emb_bio, file.path(out, "embedding_bio.csv"), row.names = FALSE)
utils::write.csv(meta, file.path(out, "cell_metadata.csv"), row.names = FALSE)
saveRDS(list(embedding = emb_bio, metadata = meta), file.path(out, "embedding_bio.rds"))
cat("saved embedding to", out, "\n")
```

## 5. Visualise with UMAP

A UMAP of the `"bio"` embedding. If integration worked, cells should group by
**cell type** (biology kept) while the **batches mix together** (batch difference
removed), colour by each to check.

```{r umap-bio, fig.width = 16, fig.height = 5}
xy <- m3_umap(emb_bio, method = "scanpy", n_neighbors = 15L, device = model$device)
df <- data.frame(UMAP1 = xy[, 1], UMAP2 = xy[, 2],
                 celltype = factor(meta$mergedcelltype),
                 batch    = factor(meta$batch),
                 condition = factor(meta$cond_group))

plt <- function(colour, title, legend = TRUE) {
  ggplot(df, aes(UMAP1, UMAP2, colour = .data[[colour]])) +
    geom_point(size = 0.4, alpha = 0.8) +
    theme_classic() +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          legend.position = if (legend) "right" else "none") +
    guides(colour = guide_legend(override.aes = list(size = 2))) +
    labs(title = title, colour = NULL)
}
print(plt("celltype",  "Cell type (biology preserved)"))
print(plt("batch",     "Batch (batch mixed)"))
print(plt("condition", "Condition"))
```

**Done.** You now have an integrated cell embedding, ready for clustering and
downstream analysis. The patient-prediction tutorial builds on the same model to
predict donor-level disease status.

```{r cleanup, include = FALSE}
m3_shutdown()
```
