---
title: "Cell-level representation learning"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(ggplot2)
```

This tutorial trains an **m3** integration model on the multi-batch, multimodal
(RNA + ADT) **Liu et al.** COVID-19 dataset and produces a batch-corrected,
condition-aware **cell embedding**.

m3's latent is **disentangled**: a cell-intrinsic part, a small per-condition
part, and a batch part. `m3_embedding(part = "bio")` returns the biology
(intrinsic + conditions, batch removed) — the representation you use for
clustering / UMAP.

```{r libs}
library(m3)
set.seed(0)
```

## 1. Read the batches and merge

Each batch is one run: an RNA HVG matrix, an ADT matrix, and a
metadata CSV. `m3_read_h5()` reads one batch into an `m3_dataset`; `m3_concat()`
stacks them, harmonising features by intersection and stamping a `batch` column.
(Adjust `DATA` to where your Liu et al files live.)

```{r data}
DATA <- Sys.getenv("M3_LIU_DATA", "/path/to/liu_et_al")
cohorts <- lapply(1:3, function(i) m3_read_h5(
  rna      = file.path(DATA, sprintf("rna%d_hvg1000.h5", i)),
  adt      = file.path(DATA, sprintf("adt%d.h5", i)),
  metadata = file.path(DATA, sprintf("metadata%d.csv", i)),
  batch    = sprintf("B%d", i)))
data <- m3_concat(cohorts)
data
```

## 2. Build and train the integration model

We declare the column roles. For *pure* representation learning we set
`donor_prediction = FALSE` so only the integration VAE is trained (the
donor-level disease predictor is Tutorial 2). `embedding_dim` is the latent
width. `batch_key` (default `"batch"`) names the obs column of batch labels
(B1/B2/B3) the VAE balances and corrects across — a real input even here.
`seed = 0` makes the (otherwise unseeded) Stage-1 VAE reproducible.

```{r train}
model <- m3_train(
  data,
  condition_keys = c("cond_group", "Age_interval"),
  celltype_key   = "mergedcelltype",
  batch_key      = "batch",
  embedding_dim  = 30L,
  donor_prediction = FALSE,
  max_epochs = 500L,
  seed = 0L
)
model
m3_capabilities(model)
```

## 3. Extract the disentangled embeddings

`part = "bio"` = cell-intrinsic + condition latents (batch removed) — the
integrated biological representation. `part = "batch"` isolates the batch latent.

```{r embed}
emb_bio       <- m3_embedding(model, part = "bio")
emb_intrinsic <- m3_embedding(model, part = "intrinsic")
emb_batch     <- m3_embedding(model, part = "batch")
meta <- m3_cell_metadata(model)
cat(sprintf("bio: %s | intrinsic: %s | batch: %s\n",
            paste(dim(emb_bio), collapse = "x"),
            paste(dim(emb_intrinsic), collapse = "x"),
            paste(dim(emb_batch), collapse = "x")))
```

## 4. Save the embedding + metadata

m3 returns plain matrices, so persist them however you like — here as CSV and an
`.rds`.

```{r save}
out <- file.path(tempdir(), "m3_tut1")
dir.create(out, showWarnings = FALSE)
utils::write.csv(emb_bio, file.path(out, "embedding_bio.csv"), row.names = FALSE)
utils::write.csv(meta, file.path(out, "cell_metadata.csv"), row.names = FALSE)
saveRDS(list(embedding = emb_bio, metadata = meta), file.path(out, "embedding_bio.rds"))
cat("saved embedding to", out, "\n")
```

## 5. Visualise with UMAP

We project the *bio* embedding with the **same** UMAP m3's Python tutorial uses
(scanpy's `neighbors` + `umap`, computed in the engine env via reticulate), so
the coordinates match exactly. Good integration = cell types form clean clusters
while batches are mixed.

```{r umap-bio, fig.width = 16, fig.height = 5}
xy <- m3_umap(emb_bio, method = "scanpy", n_neighbors = 15L, device = model$device)
df <- data.frame(UMAP1 = xy[, 1], UMAP2 = xy[, 2],
                 celltype = factor(meta$mergedcelltype),
                 batch    = factor(meta$batch),
                 condition = factor(meta$cond_group))

plt <- function(colour, title, legend = TRUE) {
  ggplot(df, aes(UMAP1, UMAP2, colour = .data[[colour]])) +
    geom_point(size = 0.4, alpha = 0.8) +
    theme_classic() +
    theme(axis.text = element_blank(), axis.ticks = element_blank(),
          legend.position = if (legend) "right" else "none") +
    guides(colour = guide_legend(override.aes = list(size = 2))) +
    labs(title = title, colour = NULL)
}
print(plt("celltype",  "Cell type (biology preserved)"))
print(plt("batch",     "Batch (batch mixed)"))
print(plt("condition", "Condition"))
```

### Batch latent (the part m3 *removed* from "bio")

Colouring a UMAP of the **batch** latent by batch shows the batch structure m3
isolates into the batch dimension and keeps out of the biological embedding.

```{r umap-batch, fig.width = 6, fig.height = 5}
xy_b <- m3_umap(emb_batch, method = "scanpy", n_neighbors = 15L, device = model$device)
ggplot(data.frame(UMAP1 = xy_b[, 1], UMAP2 = xy_b[, 2], batch = factor(meta$batch)),
       aes(UMAP1, UMAP2, colour = batch)) +
  geom_point(size = 0.4, alpha = 0.8) + theme_classic() +
  theme(axis.text = element_blank(), axis.ticks = element_blank()) +
  labs(title = "Batch latent — batch signal", colour = NULL)
```

**Done.** We trained one m3 model on three batches and obtained an integrated,
condition-aware cell embedding plus UMAPs. Tutorial 2 builds on the same model
object to predict donor-level disease status.

```{r cleanup, include = FALSE}
m3_shutdown()
```
