---
title: "Donor-level disease prediction"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(ggplot2)
```

m3's headline task: predict **donor/patient-level disease status** from multimodal
single-cell data, in a **leak-safe leave-one-batch-out** setting. We hold out one
batch (`B3`); its donors' disease labels are masked during training and
predicted at the end. One `m3_model` trains the integration VAE **and** the
donor-level adversarial predictor; `m3_predict_donors()` returns per-donor class
probabilities.

```{r libs}
library(m3)
set.seed(0)
DATA <- Sys.getenv("M3_LIU_DATA", "/path/to/liu_et_al")
cohorts <- lapply(1:3, function(i) m3_read_h5(
  rna      = file.path(DATA, sprintf("rna%d_hvg1000.h5", i)),
  adt      = file.path(DATA, sprintf("adt%d.h5", i)),
  metadata = file.path(DATA, sprintf("metadata%d.csv", i)),
  batch    = sprintf("B%d", i)))
data <- m3_concat(cohorts)
data
```

## 2. Build + train the model with a held-out batch

`held_out = "B3"` designates the query batch (leak-safe: its `cond_group`
labels are masked). `donor_key` + `celltype_key` enable the donor predictor;
`batch_key` is the site column whose effect the donor-level adversary removes;
`target_condition` is the disease axis to predict. The donor-predictor knobs are
the real engine hyperparameters used for the Liu figures.

```{r train}
model <- m3_train(
  data,
  condition_keys   = c("cond_group", "Age_interval"),
  target_condition = "cond_group",
  celltype_key     = "mergedcelltype",
  batch_key        = "batch",
  donor_key        = "sample_id",
  held_out         = "B3",
  embedding_dim    = 30L,
  max_epochs       = 500L,
  donor_predictor  = list(glr = 3e-3, n_epochs = 120L, adv_max = 10L,
                          adv_warmup = 7L, n_disc = 21L, patient_w = 10L),
  seed = 0L
)
cat("reference vocab (query labels never enter it):",
    paste(m3_reference_vocab(model)$cond_group, collapse = ", "), "\n")
```

## 3. Predict the held-out donors

```{r predict}
preds <- m3_predict_donors(model)
cat("query donors:", nrow(preds), "\n")
print(preds)
```

## 4. Evaluate against the held-out truth

The query donors' true `cond_group` lives in the metadata (never shown to the
model). We join it back to score accuracy and ROC-AUC.

```{r eval}
obs <- m3_dataset_obs(data)
truth <- unique(obs[, c("sample_id", "cond_group")])
truth <- stats::setNames(as.character(truth$cond_group), as.character(truth$sample_id))
preds$true_label <- truth[as.character(preds$donor)]
pos <- "Severe"
y_true  <- as.integer(preds$true_label == pos)
y_score <- preds[[paste0("prob_", pos)]]
acc <- mean(preds$predicted_label == preds$true_label)

auc_of <- function(score, label) {              # Mann-Whitney U / ROC-AUC
  r <- rank(score); n1 <- sum(label == 1); n0 <- sum(label == 0)
  if (n1 == 0 || n0 == 0) return(NA_real_)
  (sum(r[label == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0)
}
auc <- auc_of(y_score, y_true)
cat(sprintf("held-out accuracy = %.3f   ROC-AUC = %.3f\n", acc, auc))
```

## 5. Visualise — held-out ROC

```{r roc, fig.width = 6, fig.height = 5}
roc_points <- function(score, label) {
  o <- order(score, decreasing = TRUE)
  tp <- cumsum(label[o] == 1); fp <- cumsum(label[o] == 0)
  data.frame(fpr = c(0, fp / max(fp, 1)), tpr = c(0, tp / max(tp, 1)))
}
rp <- roc_points(y_score, y_true)
ggplot(rp, aes(fpr, tpr)) +
  geom_abline(linetype = "dashed", colour = "grey") +
  geom_line(linewidth = 1, colour = "#4c72b0") +
  annotate("text", x = 0.65, y = 0.1, label = sprintf("AUC = %.3f", auc)) +
  labs(x = "False positive rate", y = "True positive rate", title = "Held-out donor ROC") +
  theme_classic()
```

## 6. Patient-level embedding

`m3_donor_embedding()` returns the **patient-level (donor) embedding** — the
corrected donor vector the model actually classifies (one row per donor). We
UMAP it (same `umap-learn`, `random_state = 0`, as the Python tutorial) and
colour by reference/query, true phenotype, and whether the held-out prediction
was correct.

```{r patient-umap, fig.width = 16, fig.height = 5}
demb <- m3_donor_embedding(model)
info <- m3_predict_donors(model, include_reference = TRUE)
info <- info[match(demb$donor, info$donor), ]
info$true_label <- truth[as.character(demb$donor)]
info$set     <- ifelse(info$is_reference, "reference", "query")
info$correct <- ifelse(info$is_reference, "reference",
                       ifelse(info$predicted_label == info$true_label, "correct", "wrong"))

X  <- as.matrix(demb[, grep("^m3_", colnames(demb))])
xy <- m3_umap(X, method = "umap", n_neighbors = 15L, random_state = 0L,
              device = model$device)
pdf <- data.frame(UMAP1 = xy[, 1], UMAP2 = xy[, 2],
                  set = factor(info$set), true_label = factor(info$true_label),
                  correct = factor(info$correct))

pl <- function(colour, title) {
  ggplot(pdf, aes(UMAP1, UMAP2, colour = .data[[colour]])) +
    geom_point(size = 3, alpha = 0.85) + theme_classic() +
    theme(axis.text = element_blank(), axis.ticks = element_blank()) +
    labs(title = title, colour = NULL)
}
print(pl("set",        "Patient embedding — set"))
print(pl("true_label", "Patient embedding — true_label"))
print(pl("correct",    "Patient embedding — correct"))
```

**Done.** Leak-safe leave-one-batch-out donor disease prediction with a held-out
ROC and a patient-level embedding UMAP.

```{r cleanup, include = FALSE}
m3_shutdown()
```
