---
title: "Feature attribution"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(ggplot2)
```

Once m3 can predict disease, attribution asks which genes, cells and cell types
drove that prediction. We train on the full reference, then trace each prediction
back to the genes, cells, and cell types behind it.

## 1. Read the batches and merge

```{r libs}
library(m3)
set.seed(0)
DATA <- Sys.getenv("M3_LIU_DATA", "/path/to/liu_et_al")
cohorts <- lapply(1:3, function(i) m3_read_h5(
  rna      = file.path(DATA, sprintf("rna%d_hvg1000.h5", i)),
  adt      = file.path(DATA, sprintf("adt%d.h5", i)),
  metadata = file.path(DATA, sprintf("metadata%d.csv", i)),
  batch    = sprintf("B%d", i)))
data <- m3_concat(cohorts)
data
```

## 2. Build and train the model

Same setup as the patient-prediction tutorial, minus the held-out batch (we
attribute on the full reference). `m3_train(...)` fits both the integration model
and the disease predictor; attribution explains that predictor.

```{r train}
model <- m3_train(
  data,
  condition_keys   = c("cond_group", "Age_interval"),
  target_condition = "cond_group",
  celltype_key     = "mergedcelltype",
  batch_key        = "batch",
  donor_key        = "sample_id",
  embedding_dim    = 30L,
  max_epochs       = 500L,
  seed = 0L
)
m3_capabilities(model)
```

## 3. Attribute and rank the cell types

`m3_attribute(model, reference_labels = "HC")` scores how much each gene, cell,
and cell type pushed the prediction away from the reference label (HC, i.e.
healthy). Here we look at the **top cell types**: `m3_top_celltypes(...)` ranks
them, keeping only cell types with enough cells in each condition so a small group
can't mislead the ranking.

```{r attribute}
attr <- m3_attribute(model, reference_labels = "HC")   # n_steps defaults to 50
cat("\ntop cell types (>= 200 cells per condition):\n")
print(m3_top_celltypes(attr, min_cells_per_condition = 200L))
```

## 4. Top genes

`m3_top_genes(...)` ranks the genes by how strongly they drove the prediction.
`min_cells_per_condition` keeps only cell types with enough cells in each condition
(so a tiny group can't dominate the ranking), and housekeeping / ribosomal genes
are dropped.

```{r top-genes}
top100_rna <- m3_top_genes(attr, n = 100L, min_cells_per_condition = 200L, modality = "rna")
cat(sprintf("top-100 RNA genes (computed over %d balanced cell types):\n",
            top100_rna$n_celltypes_used[1]))
print(utils::head(top100_rna, 15))
```

**Done.** The genes, cells, and cell types behind m3's Severe-vs-HC prediction.

```{r cleanup, include = FALSE}
m3_shutdown()
```
