---
title: "Donor-level disease prediction"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
library(ggplot2)
```

m3 can predict a **donor's** disease status from their cells, in a
leave-one-batch-out setting: we hold out one batch (`B3`), hide its donors'
disease labels during training, and predict them at the end. `m3_train(...)` fits
the integration model and the donor predictor in one call; `m3_predict_donors()`
then returns per-donor class probabilities.

## 1. Load the demo dataset

```{r libs}
library(m3)
set.seed(0)
data <- m3_demo()
data
```

## 2. Build + train the model with a held-out batch

On top of the columns from the representation-learning tutorial, donor prediction
needs a few more:

- `target_condition`, the condition to predict (here `cond_group`, i.e. disease).
- `donor_key`, the column identifying each donor, so cells are grouped per donor.
- `held_out`, the batch(es) to hold back; these donors' `target_condition` labels
  are hidden during training and predicted at the end.

The `donor_predictor` list holds the predictor's training settings (learning rate,
epochs, and how strongly / on what schedule the batch effect is removed).
`m3_reference_vocab(model)` shows the label set the model learned to predict (the
held-out donors' labels never enter it).

```{r train}
model <- m3_train(
  data,
  condition_keys   = c("cond_group", "Age_interval"),
  target_condition = "cond_group",
  celltype_key     = "mergedcelltype",
  batch_key        = "batch",
  donor_key        = "sample_id",
  held_out         = "B3",
  embedding_dim    = 30L,
  max_epochs       = 500L,
  donor_predictor  = list(glr = 3e-3, n_epochs = 120L, adv_max = 10L,
                          adv_warmup = 7L, n_disc = 21L, patient_w = 10L),
  seed = 0L
)
cat("reference vocab (query labels never enter it):",
    paste(m3_reference_vocab(model)$cond_group, collapse = ", "), "\n")
```

## 3. Predict the held-out donors

`m3_predict_donors(model)` returns one row per held-out donor: the predicted label
and the probability of each class.

```{r predict}
preds <- m3_predict_donors(model)
cat("query donors:", nrow(preds), "\n")
print(preds)
```

## 4. Evaluate against the held-out truth

The held-out donors' real `cond_group` lives in the metadata (never shown to the
model). We join it back to score accuracy.

```{r eval}
obs <- m3_dataset_obs(data)
truth <- unique(obs[, c("sample_id", "cond_group")])
truth <- stats::setNames(as.character(truth$cond_group), as.character(truth$sample_id))
preds$true_label <- truth[as.character(preds$sample_id)]
acc <- mean(preds$predicted_label == preds$true_label)
cat(sprintf("held-out accuracy = %.3f\n", acc))
```

## 5. Patient-level embedding

`m3_donor_embedding()` returns one vector per donor, the donor-level
representation the model actually classifies. We UMAP it and colour by
reference/query, true label, and whether the held-out prediction was correct.

```{r patient-umap, fig.width = 16, fig.height = 5}
demb <- m3_donor_embedding(model)
info <- m3_predict_donors(model, include_reference = TRUE)
info <- info[match(demb$sample_id, info$sample_id), ]
info$true_label <- truth[as.character(demb$sample_id)]
info$set     <- ifelse(info$is_reference, "reference", "query")
info$correct <- ifelse(info$is_reference, "reference",
                       ifelse(info$predicted_label == info$true_label, "correct", "wrong"))

X  <- as.matrix(demb[, grep("^m3_", colnames(demb))])
xy <- m3_umap(X, method = "umap", n_neighbors = 15L, random_state = 0L,
              device = model$device)
pdf <- data.frame(UMAP1 = xy[, 1], UMAP2 = xy[, 2],
                  set = factor(info$set), true_label = factor(info$true_label),
                  correct = factor(info$correct))

pl <- function(colour, title) {
  ggplot(pdf, aes(UMAP1, UMAP2, colour = .data[[colour]])) +
    geom_point(size = 3, alpha = 0.85) + theme_classic() +
    theme(axis.text = element_blank(), axis.ticks = element_blank()) +
    labs(title = title, colour = NULL)
}
print(pl("set",        "Patient embedding — set"))
print(pl("true_label", "Patient embedding — true_label"))
print(pl("correct",    "Patient embedding — correct"))
```

**Done.** Leave-one-batch-out donor prediction, evaluated against the held-out
truth, with a patient-level embedding.

```{r cleanup, include = FALSE}
m3_shutdown()
```
